Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
add Lunt2021 to table

data <- data.frame(
  Source = c("Bijl et al. (2009)", "Keating-Bitoni et al. (2011)", "Tierney et al. (2017)", "Cramwinckel et al. (2018)", "Evans et al. (2018)", "Pross et al. (2012), as shown in Tierney et al. (2017)"),
  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE"),
  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26"),
  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle"),
  Model = c("$2^{nd}$ order polynomial",
            "$2^{nd}$ order polynomial",
            "Gaussian function",
            "-",
            "-",
            "climate model ensemble (GCM)"),
  Proxy_system = c("$TEX_{86}$, $UK_{37}^{K'}$",
                   "$TEX_{86}$, MBT/CBT, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$",
                   "$TEX_{86}$",
                   "$TEX_{86}$, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$, deepwater $\\delta^{18}O$",
                   "$\\Delta_{47}$, deepwater Mg/Ca",
                   "none (GCM simulations)")
)

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

>>>>

data <- data.frame(
  Source = c("Bijl et al. (2009)", "Keating-Bitoni et al. (2011)", "Tierney et al. (2017)", "Cramwinckel et al. (2018)", "Evans et al. (2018)", "Pross et al. (2012), as shown in Tierney et al. (2017)", "Lunt et al. (2021)"),
  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE", "EECO"),
  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26", "18 - 26"), 
  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle", "tropics - high latitude"),
  Model = c("$2^{nd}$ order polynomial",
            "$2^{nd}$ order polynomial",
            "Gaussian function",
            "-",
            "-",
            "climate model ensemble",
            "DeepMIP climate models"),
  Proxy_system = c("$TEX_{86}$, $UK_{37}^{K'}$",
                   "$TEX_{86}$, MBT/CBT, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$",
                   "$TEX_{86}$",
                   "$TEX_{86}$, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$, deepwater $\\delta^{18}O$",
                   "$\\Delta_{47}$, deepwater Mg/Ca",
                   "none (GCM simulations)",
                   "none (GCM simulations)")
)

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
